The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-10 paper/iphas-dr2.tex

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against the Sloan Digital Sky Survey (SDSS),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure 
of our own backyard, the Milky Way.
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the INT in La Palma --
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}
to be extended into larger, deeper, and more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

@@

A new data release for the INT/WFC Photometric \ha\ Survey
of the Northern Galactic Plane (IPHAS) was presented.
It is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
taking in the entire Northern Milky Way at $|b|<5^\circ$.
The new 97-column catalogue provides single-epoch photometry
across 92\% of the survey area,
and is the first quality-controlled and uniformly calibrated
catalogue to have been constructed from the imaging data.
We explained the data reduction and quality control procedures that
were applied,
described and tested the global photometric calibration,
and detailed the construction of the source catalogue.

The observations included in this release
achieved a median seeing of 1.1 arcsec
and a $5\sigma-$depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$, \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
were found to be accurate at the level of $0.03$ mag (rms),
providing a significant improvement over the 
previous data release.
The source catalogue provides the best-available
single-epoch astrometry and photometry
for 219~million unique sources.
We detailed a list of recommended quality criteria
to select objects with reliable colours from the catalogue.
Finally, we demonstrated the use of the survey's unique
$(r-\ha,\ r-i)$ diagram for characterising stellar populations
and selecting emission-line objects.

In future work, we intend to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We also aim to correct the IPHAS photometry for the
radial field distortions,
plan to produce a catalogue 
which details all the secondary detections
to aid time-domain studies,
and are investigating the use of photometry
derived from PSF-fitting across the survey.

The data-taking strategy developed for IPHAS
has since been reapplied to carry out 
a companion INT/WFC Galactic Plane survey called UVEX
in U, $g$, $r$, He {\sc i} \citep{Groot2009},
a survey of the Kepler field 
in $U$, $g$, $r$, $i$, \ha\ 
\citep{Greiss2012},
and a survey of the entire Southern Galactic Plane
in $u$, $g$, $r$, $i$, \ha\ 
called VPHAS+ \citep{Drew2014}.
We intend to use the present work
as a template for generating catalogues
from these companion surveys.
When combined, these surveys
will offer new insights
into the contents and structure of the Milky Way
across 360$^\circ$,
and provide images across the most crowded
regions to be faced by Gaia.

>>>>

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering the Galactic latitude range $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against the Sloan Digital Sky Survey (SDSS),
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the INT in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

@@

A new data release for the INT/WFC Photometric \ha\ Survey
of the Northern Galactic Plane (IPHAS) was presented.
It is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
taking in the entire Northern Milky Way at $|b|<5^\circ$.
The new 97-column catalogue provides single-epoch photometry
across 92\% of the survey area,
and is the first quality-controlled and uniformly calibrated
catalogue to have been constructed from the imaging data.
%We explain the data reduction and quality control procedures that
%were applied.  We describe and test the global photometric 
%calibration, 
%and detailed the construction of the source catalogue.

The observations included in this release
achieve a median seeing of 1.1 arcsec
and a $5\sigma-$depth of $r=21.2\pm 0.5$, $i=20.0\pm 0.3$, \ha$=20.3\pm 0.3$.
The global calibration and photometric repeatability
are found to be accurate at the level of $0.03$ mag (rms),
providing a significant improvement over the 
previous data release.
The source catalogue specifies the best-available
single-epoch astrometry and photometry
for 219~million unique sources.
To support its exploitation, we provide a list of recommended quality criteria
that will permit the selection of objects with reliable colours from 
the catalogue.  The closing demonstrations highlight the use of the 
survey's unique $(r-\ha,\ r-i)$ diagram for characterising stellar populations
and selecting emission-line objects.  More comprehensive applications of IPHAS
can be found in the works of Sale et al (2014), which applies DR2 to the problem
of 3D extinction mapping, and of Sabin et al (2014), where the results of a search
of the image database for new planetary nebulae is presented.

The current plan is to work toward one further major IPHAS source catalogue, in which 
the remaining gaps in sky coverage will have been eliminated -- observations aimed at 
replacing data not meeting the IPHAS minimum quality requirements continue.  We will also 
examine options to further improve the global calibration, perhaps tightening the accuracy to
better than 2\%.  For example, we will investigate the use of e.g. the PanSTARRS photometric 
ladder (Magnier et al 2013) as a reference set, and explore improving source recovery in 
the most dense fields via the implementation of PSF fitting in place of aperture photometry.  
Finally, the next catalogue will detail all the secondary detections to aid time-domain studies.

The data-taking strategy developed for IPHAS
has since been reapplied to carry out 
a companion INT/WFC Galactic Plane survey called UVEX
in U, $g$, $r$, He {\sc i} \citep{Groot2009},
a survey of the Kepler field 
in $U$, $g$, $r$, $i$, \ha\ 
\citep{Greiss2012},
and a survey of the Southern Galactic Plane and Bulge
in $u$, $g$, $r$, $i$, \ha\ 
called VPHAS+ \citep{Drew2014}.  The work presented here
stands as a potential template for the catalogues that
remain to be generated from these sibling surveys.  
In prospect from them, whether they are mined separately or together,
are the means to ask seamless questions on the contents and structure of 
the most highly-populated components of the Milky Way.
